---
title: "Intro to Clinical Data: Part I"
author: Nicole Black, Kayla Xu
date: 07/17/22
updated: 02/12/23
---

Welcome to your first experience using TCGA data in R! The first data type we will begin with is clinical data (ex: sex, age at initial diagnosis, tumor subtype, etc.)

###### (1) Set up your working environment

Before running this code, make sure you are in the correct working directory (your personal qbio_490_name folder) with getwd() and setwd(). Otherwise you’ll have trouble tracking the data down!

As good general practice to keep things organized, create an "analysis_data" folder in the root of your GitHub repository (your qbio_490_name folder), then set that folder as your working directory ("/PATH/TO/qbio_490_name/analysis_data") whenever you are working with TCGA files. You can create the folder through your command line now.

Once the folder is created, set your working directory below:

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/aly/Desktop/Main/School_Files/S2/QBIO/qbio_490_alexzhang/analysis_data"))
getwd()
```

Before you move on, go to your GitHub web browser, open your .gitignore, and make sure the following is added in. 

      "### TCGA Data
      GDCdata
      MANIFEST*
      htseq_h5*
      analysis_data"

The loaded files from TCGA are quite large, so we don't want to push them to GitHub, it's also typically best to avoid putting raw data on GitHub.

#checked and everything is added in .gitignore already

###### (2) Query and retrieve the TCGA breast cancer clinical data 

```{r}
clin_query <- GDCquery(project = "TCGA-BRCA",
                      data.category = "Clinical",
                      file.type = "xml")
  
clinic <- GDCprepare_clinic(clin_query,
                            clinical.info = "patient")

```

Querying TCGA data can take a while, even without downloading. A faster way to save and access the clinical data is to save the dataframe as a csv file on your local computer. Use the following command to do so:

```{r}
#datanull <- NULL
#write.csv(datanull, file='brca_clinical_data.csv', row.names=FALSE)
#testing to see if full path was needed, as the full path didnt work

write.csv(clinic, file='brca_clinical_data.csv', row.names=FALSE)
```

To read the csv back into your environment as a dataframe:

```{r}
#full path messes up my r idk why but just filename works. probs bc i set working directory...?
clinic <- read.csv('brca_clinical_data.csv')
```



###### (3) Exploring the Data

View the structure of the data as well as the first few lines. What two commands should you use?
Answer the following questions based on the output of those commands: 
1. How many rows are there? How many columns?
2. What is the data type of $gender?
3. What is the age in years of the 5th patient (barcode: TCGA-4H-AAAK)?
4. How old was the 1st patient at their last follow up (barcode: TCGA-3C-AAAU)? HINT: you need to look at data from two columns to answer this

```{r Exercise 3.1}
#1
nrow(clinic)
ncol(clinic)
#1174 row, 114 column

#2
data.class('gender')
#character

#3
#there is only age_at_diagnosis and days_to_birthday
#ageof5th = clinic[clinic$barcode == 'TCGA-4H-AAAK', 'age']
#ageof5th <- as.integer(ageof5th)
#is.integer(ageof5th)
#print(ageof5th)
print(head(clinic))
days_to_birth_5th = 18371/365
age_in_years_of_5th <- days_to_birth_5th
age_in_years_of_5th
#50.331 years old

#4
install.packages("dplyr")
install.packages("tidyr")
library(dplyr)
library(tidyr)
patient1 <- read.csv('brca_clinical_data.csv') %>%
  dplyr::select(days_to_birth, days_to_last_followup) %>%
  extract(days_to_birth, into = "numYR") %>%
  extract(days_to_last_followup, into = "numFL") %>%
  dplyr::mutate(as.numeric(numYR)) %>%
  dplyr::mutate(as.numeric(numFL))
#run this above together
  
  patient1$numYR
  patient1$numFL
#run this to see years in days
ageAtLV = (20211-3767)/365
#age of patient at last visit is 45.05 years

#couldn't remember how to do it the easy way until i read the next set of instructions :(

```

View the column and row names. What two commands should you use?

```{r Exercise 3.2}
colnames(clinic)
rownames(clinic)
```

View the vital status column. What two notations can be used to view columns? 

```{r Exercise 3.3}

clinic$vital_status
clinic['vital_status']

```

###### (4) Plotting

Choose two variables and make a scatter plot (recall the plot() function).

THINK: what types of variables can you use in a scatter plot (continuous, discrete, categorical). Look these up if you need help!

For an extra challenge, add a plot title, axes titles, and colors to your plot.

```{r Exercise 4.1}
patient1
plot(patient1$numYR, patient1$numFL, type = "p", main = "Age Compared to Time Since Last Diagnosis", xlim = c(0, 30000), ylim = c(0, 10000), xlab = "Age", ylab = "Time Since Last Diagnosis")
```

Choose two new variables and make a box plot (recall the boxplot() function).

THINK: what types of variables can you use in a box plot?

For an extra challenge, add a plot title, axes titles, and confidence intervals to your plot.

```{r Exercise 4.2}
#clinic
numclinic <- read.csv('brca_clinical_data.csv') %>%
  dplyr::select(age_at_initial_pathologic_diagnosis, year_of_initial_pathologic_diagnosis) %>%
  extract(age_at_initial_pathological_diagnosis, into = "agePathDiag") %>%
  extract(year_of_initial_pathologic_diagnosis, into = "yearPathDiag") %>%
  dplyr::mutate(as.numeric(agePathDiag)) %>%
  dplyr::mutate(as.numeric(yearPathDiag))

  #plot doesn't work unless run separately idek why
  #converted variables with as.numeric(), but when i go to run the code chunk it doesn't work
  #HOWEVER, when i go to run the code separately (the parts above and below this comment), IT WORKS SOMEHOW, THE PLOT IS THERE(?)
  
plot(numclinic$age_at_initial_pathologic_diagnosis, numclinic$year_of_initial_pathologic_diagnosis, type = "b", main = "age at initial pathological diagnosis and year of initial pathological diagnosis", xlim = c(0, 100), ylim = c(1500, 2500), xlab = "age at initial pathological diagnosis", ylab = "year of initial pathological diagnosis")
```


###### (5) Boolean Masking

As discussed in lecture, a boolean mask is a vector of True and False values that allow for fast and efficient indexing of large dataframes in R. Boolean masks work as an extremely helpful implementation of R's vectorization abilities. 

A significant portion of your coding work will be data cleaning and preparation using boolean masks. As such, it's important you are comfortable with making them.

1) Create a mask to remove all NA values from vital_status
```{r}
vital_status_na_mask <- !is.na(clinic$vital_status)
clinic$vital_status[vital_status_na_mask]
```

2) Create a mask to select for male patients
```{r}
clinic$gender
#male called "MALE"
male_mask <- clinic$gender == "MALE"
clinic$gender[male_mask]
```

3) Create a mask to select for patients under the age of 35 OR over 60 at diagnosis. 
```{r}
under30_over60_mask <- (clinic$age_at_initial_pathologic_diagnosis < 35)|(clinic$age_at_initial_pathologic_diagnosis > 60)
clinic$age_at_initial_pathologic_diagnosis[under30_over60_mask]
```

4) Using your under30_over60_mask, create a new dataframe called no_middle_clinic which contains all the clinical data for patients under the age of 30 or over 60 at diagnosis

```{r}
no_middle_clinic <- clinic[under30_over60_mask,]
```


###### (6) Submitting Coding HW

Your homework should be a secure copy (scp) of the original file in sp23_course_materials. Rename your file to include your name at the end (ie. Intro_to_Clinical_Data_I_kayla.Rmd). Push the file to Github through your qbio_490_name local repository and submit a link to your github repo to Blackboard. 

Now you can move on to part II!
